Computing solution space properties of combinatorial optimization problems via generic tensor networks

### 2022-08-29 paper/paper.tex
fix a theorem
Updates from Overleaf

\section{Tensor network representation of independent sets} \label{sec:tnmap}
This section describes the reduction of the independent set problem to a tensor network contraction problem.
An alternative interpretation, perhaps more accessible to physicists, can be found in \Cref{sec:energymodel}, where we introduce the reduction from the energy model of hardcore lattice gases~\cite{Dyre2016, Fernandes2007}.
Let $G=(V,E)$ be an undirected graph with each vertex $v\in V$ being associated with a weight $w_v$.
An independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, we have $(u, v) \not\in E$.
To reduce the independent set problem on $G$ to a tensor network contraction, we first map each vertex $v\in V$ to a label $s_v \in \{0, 1\}$ of dimension $2$, where we use $0$ ($1$) to denote a vertex is absent (present) in $I$.
For each label $s_v$, we define a parameterized rank-one tensor associated with it as
\begin{equation}
    W^{(v)} = \left(\begin{matrix}
        1 \\
        x_v^{w_v}
    \end{matrix}\right),\label{eq:vertextensor}
\end{equation}
where $x_v$ is a variable associated with $v$.
Tensor elements can be indexed by $s_v$, e.g.\ $W^{(v)}_0=1$ is the first element associated with $s_v=0$ and $W^{(v)}_1=x_v^{w_v}$ is the second element associated with $s_v=1$.
Similarly, for each edge $(u, v) \in E$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B^{(u, v)} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right), \label{eq:edgetensor}
\end{equation}
which encodes the independent set constraint since $B_{11} = 0$, meaning the two vertices cannot be both in the independent set. 
The corresponding tensor network $\mathcal{N}_{\rm IS}(G)$ is
\begin{equation}\label{eq:mistensornetwork}
\begin{split}
    \mathcal{T} &= \{W^{(v)} | v\in V\} \cup \{B^{(u, v)} | (u, v) \in E\} \cup \{P\}\\
    \Sigma &= \{s_v | v \in V\}\\
    O &= P\\
    \lambda &= T \mapsto \begin{cases}
                    (s_v), &T = W^{(v)}\\
                    (s_u, s_v), &T = B^{(u, v)}\\
                    (), & T = P
                    \end{cases}
    \end{split}
\end{equation}
where we use the superscripts to denote which vertex or edge a tensor is associated to.
$P$ is the output tensor of rank $0$, i.e. a scalar.
The contraction of this tensor network is
\begin{equation}\label{eq:idp}
    P(G) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{v\in V} W^{(v)}_{s_v} \prod\limits_{(u,v) \in E} B^{(u,v)}_{s_u s_v},
\end{equation}
where the summation runs over all $2^{|V|}$ vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$. 
% It is easy to see that an edge tensor element $B_{s_us_v}$ encodes the independent set constraint. It zeros out the product whenever both $u$ and $v$ are in the independent set.
A vertex tensor element $W^{(v)}_{s_v}$ contributes a multiplicative factor $x_v^{w_v}$ whenever $v$ is in the set.

@@

\begin{theorem}\label{thm:complexpoly}
    The independence polynomial~\cite{Harvey2018,Ferrin2014} of a graph $G = (V, E)$ can be computed in time $O(|V|\log(|V|))\cc$.
\end{theorem}
\begin{proof}
By doing the following replacement of tensor elements from the standard number type to the polynomial number type
\begin{equation}
    \begin{cases}
    1 \rightarrow \mymathbb{1}\\
    0 \rightarrow \mymathbb{0}\\
    x_v^{w_v} \rightarrow (0, 1)
    \end{cases}
\end{equation}
The tensors $W$ and $B$, which are introduced in the previous section (\Cref{sec:tnmap}), can thus be written as 
\begin{equation}
    \left(W^{\rm PN}\right)^{(v)} = \left(\begin{matrix}
        \mymathbb{1} \\
        (0,1)
    \end{matrix}\right),   
    \qquad \qquad
        \left(B^{\rm PN}\right)^{(u, v)} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
By contracting the tensor network with this polynomial type, we have the exact representation of the independence polynomial.
In \Cref{eq:polynomial}, the addition can be computed in time $O(k)$, where $k=\max(k_a, k_b)$, and the multiplication can be evaluated in time $O(k\log(k))$ using the convolution theorem~\cite{Schonhage1971}.
Since $k$ is upper bounded by the maximum independent set size $\alpha(G) \leq |V|$, the time complexity of element-wise addition and multiplication operation is upper bounded by $|V|\log(|V|)$. Combining with \Cref{thm:complexreal}, the a overall time complexity is $O(|V|\log(|V|))\cc$.
\end{proof}

@@

\begin{theorem}\label{thm:complexset}
    Let $G = (V, E)$ be a graph and $I$ be its independent sets.
    $I$ can be computed in time $O(|I||E|)\cc$.
\end{theorem}
\begin{proof}
To enumerate over $I$,  we initialize variable $x_{i}$ in the vertex tensor to $x_i = \{\boldsymbol{e}_{i}\}$, where $\boldsymbol{e}_i$ is a basis bit string of size $|V|$ that has only one non-zero value at location $i$.
The vertex and edge tensors are thus
\begin{equation}
    \left(W^{\rm SN}\right)^{(v)} = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}
    \end{matrix}\right),   
    \qquad \qquad
        \left(B^{\rm SN}\right)^{(u, v)} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The contraction of this tensor network is a set with its elements being all independent set configurations. The time complexity of element-wise addition and multiplication is upper bounded by $|I||V|$, where $|V|$ comes from the linear cost of representing a $|V|$-vertex configuration. Combining with \Cref{thm:complexreal}, the overall time complexity is $O(|I||V|)\cc$.
In practise, the huge multiplicative factor $|I|$ only appear in the last step of tensor contraction.
This complexity only upper bounds the actually performance of the algorithm.
\end{proof}

>>>>

\section{Tensor network representation of independent sets} \label{sec:tnmap}
This section describes the reduction of the independent set problem to a tensor network contraction problem.
An alternative interpretation, perhaps more accessible to physicists, can be found in \Cref{sec:energymodel}, where we introduce the reduction from the energy model of hardcore lattice gases~\cite{Dyre2016, Fernandes2007}.
Let $G=(V,E)$ be an undirected graph with each vertex $v\in V$ being associated with a weight $w_v$.
An independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, we have $(u, v) \not\in E$.
To the independent set problem on $G$ can be encoded as a tensor network $\mathcal{N}_{\rm IS}(G)$
\begin{equation}\label{eq:mistensornetwork}
\begin{split}
    \mathcal{T} &= \{W^{(v)} | v\in V\} \cup \{B^{(u, v)} | (u, v) \in E\} \cup \{P\},\\
    \Sigma &= \{s_v | v \in V\},\\
    O &= P,\\
    \lambda &= T \mapsto \begin{cases}
                    (s_v), &T = W^{(v)}\\
                    (s_u, s_v), &T = B^{(u, v)}\\
                    (), & T = P
                    \end{cases},
    \end{split}
\end{equation}
where for each vertex $v$, we define a parameterized rank-one tensor associated with it as
\begin{equation}
    W^{(v)} = \left(\begin{matrix}
        1 \\
        x_v^{w_v}
    \end{matrix}\right),\label{eq:vertextensor}
\end{equation}
and for each edge $(u, v) \in E$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B^{(u, v)} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
We map each vertex $v\in V$ to a label $s_v \in \{0, 1\}$ of dimension $2$, where we use $0$ ($1$) to denote a vertex is absent (present) in $I$.
%We use the superscripts to denote which vertex or edge a tensor is associated to.
These labels can be used as subscripts of tensors to index tensor elements, e.g.\ $W^{(v)}_0=1$ is the first element associated with $s_v=0$ and $W^{(v)}_1=x_v^{w_v}$ is the second element associated with $s_v=1$, where $x_v$ is a variable associated with vertex $v$.
$P$ is the output tensor of rank $0$, i.e. a scalar.
The independence constraint is encoded in the edge tensors, where we use $B_{11} = 0$ to denote two vertices connected by an edge cannot be both in the independent set. 
The contraction of this tensor network is
\begin{equation}\label{eq:idp}
    P(G) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{v\in V} W^{(v)}_{s_v} \prod\limits_{(u,v) \in E} B^{(u,v)}_{s_u s_v},
\end{equation}
where the summation runs over all $2^{|V|}$ vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$. 
% It is easy to see that an edge tensor element $B_{s_us_v}$ encodes the independent set constraint. It zeros out the product whenever both $u$ and $v$ are in the independent set.
A vertex tensor element $W^{(v)}_{s_v}$ contributes a multiplicative factor $x_v^{w_v}$ whenever $v$ is in the set.

@@

\begin{theorem}\label{thm:complexpoly}
    The independence polynomial~\cite{Harvey2018,Ferrin2014} of a graph $G = (V, E)$ can be computed in time $O(|V|\log(|V|))\cc$.
\end{theorem}
\begin{proof}
By doing the following replacement of tensor elements from the standard number type to the polynomial number type
\begin{equation}
    \begin{cases}
    1 &\rightarrow \mymathbb{1},\\
    0 &\rightarrow \mymathbb{0},\\
    x_v^{w_v} &\rightarrow (0, 1),
    \end{cases}
\end{equation}
the tensors $W$ and $B$, which are introduced in the previous section (\Cref{sec:tnmap}), can thus be written as 
\begin{equation}
    \left(W^{\rm PN}\right)^{(v)} = \left(\begin{matrix}
        \mymathbb{1} \\
        (0,1)
    \end{matrix}\right),   
    \qquad \qquad
        \left(B^{\rm PN}\right)^{(u, v)} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
By contracting the tensor network with this polynomial type, we have the exact representation of the independence polynomial.
In \Cref{eq:polynomial}, the addition can be computed in time $O(k)$, where $k=\max(k_a, k_b)$, and the multiplication can be evaluated in time $O(k\log(k))$ using the convolution theorem~\cite{Schonhage1971}.
Since $k$ is upper bounded by the maximum independent set size $\alpha(G) \leq |V|$, the time complexity of element-wise addition and multiplication operation is upper bounded by $|V|\log(|V|)$. Combining with \Cref{thm:complexreal}, the a overall time complexity is $O(|V|\log(|V|))\cc$.
\end{proof}

@@

\begin{theorem}\label{thm:complexset}
    The independent sets of graph $G = (V, E)$ be a graph can be enumerated in time $O(|I|)\cc$, where $|I|$ is the number of independent sets.
\end{theorem}
\begin{proof}
To enumerate over $I$,  we initialize variable $x_{i}$ in the vertex tensor to $x_i = \{\boldsymbol{e}_{i}\}$, where $\boldsymbol{e}_i$ is a basis bit string of size $|V|$ that has only one non-zero value at location $i$.
The vertex and edge tensors are thus
\begin{equation}
    \left(W^{\rm SN}\right)^{(v)} = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}
    \end{matrix}\right),   
    \qquad \qquad
        \left(B^{\rm SN}\right)^{(u, v)} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The contraction of this tensor network is a set with its elements being all independent set configurations. The time complexity of element-wise addition and multiplication is upper bounded by $|I||V|$, where $|V|$ comes from the linear cost of representing a $|V|$-vertex configuration. Combining with \Cref{thm:complexreal}, the overall time complexity is $O(|I||V|)\cc$.
In practise, the huge multiplicative factor $|I|$ only appear in the last step of tensor contraction.
This complexity only upper bounds the actually performance of the algorithm.
\end{proof}
